\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{enumitem}
\usepackage{graphicx}
\usepackage{array}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning}

\title{Wasm to NeoVM Translation Specification}
\author{R3E Network}
\date{\today}

\begin{document}
\setlength{\emergencystretch}{2em}
\sloppy
\hfuzz=2pt
\hbadness=10000
\maketitle
\tableofcontents
\newpage

\section{Scope}
This specification defines the normative behaviour of the \texttt{wasm-neovm} translator that lowers WebAssembly (Wasm) modules into NeoVM scripts and emits the accompanying NEF and manifest artefacts. It covers:
\begin{itemize}[nosep]
    \item Acceptance rules governing Wasm structures, imports/exports, custom sections, and unsupported features.
    \item A complete Wasm instruction to NeoVM opcode mapping, including runtime helpers for memory, tables, and reference types.
    \item Manifest generation, overlay semantics, safe-flag propagation, feature inference, and the CLI diff workflow.
    \item NEF emission, method-token ingestion, source metadata, and checksum rules.
\end{itemize}

\section{Definitions}
\begin{description}[style=unboxed,leftmargin=0cm]
    \item[Wasm module] A binary payload conforming to the WebAssembly Core Specification.
    \item[NeoVM script] A byte sequence of NeoVM opcodes produced by the translator; it is embedded into the NEF artefact unchanged.
    \item[Manifest overlay] A JSON fragment (embedded via \texttt{neo.manifest} or provided externally) that is deep-merged into the generated manifest.
    \item[Translation configuration] The tuple of contract name, optional overlays, and CLI overrides supplied to the translator.
\end{description}

\section{Pipeline Overview}
\subsection{Stages}
\begin{enumerate}[label=\arabic*., nosep]
    \item \textbf{Frontend}: Parse sections, register type signatures, imports, exports, tables, globals, data/element segments, and custom sections.
    \item \textbf{Translation}: Lower Wasm operators to NeoVM opcodes, synthesise runtime helpers lazily, and accumulate manifest metadata.
    \item \textbf{Manifest Assembly}: Build a baseline manifest from translated exports, merge overlays, propagate safe flags, infer features, and enforce ABI parity.
    \item \textbf{NEF Emission}: Serialize the script + metadata into a NEF file, including compiler ID, method tokens, optional source URL, and checksum.
\end{enumerate}

\subsection{Data Flow}
\begin{center}
\begin{tabular}{>{\raggedright}p{3.6cm} >{\raggedright}p{5cm} >{\raggedright\arraybackslash}p{4.4cm}}
\toprule
Stage & Inputs & Outputs \\
\midrule
Frontend & Wasm bytes, \texttt{ModuleTypes} accumulator & IR: types, imports, exports, tables, globals, data/element segments, custom sections \\
Translation & IR, \texttt{RuntimeHelpers} & NeoVM script bytes, method metadata (name, params, return, offset), inferred method tokens, optional source \\
Manifest & Methods, overlays & Validated manifest JSON with safe flags and inferred features \\
NEF Writer & Script, metadata & \texttt{*.nef} + \texttt{*.manifest.json} artefacts \\
\bottomrule
\end{tabular}
\end{center}
Figure~\ref{fig:pipeline} illustrates the stages and optional inputs at a glance.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[>=Stealth, node distance=2cm, every node/.style={font=\small}]
\tikzstyle{stage}=[draw, rounded corners, fill=gray!10, minimum width=3.8cm, minimum height=1.1cm]
\node[stage] (frontend) {\shortstack{1. Frontend\\{\footnotesize Parse Wasm sections}}};
\node[stage, right=1.8cm of frontend] (translator) {\shortstack{2. Translation\\{\footnotesize Emit NeoVM + helpers}}};
\node[stage, right=1.8cm of translator] (manifest) {\shortstack{3. Manifest\\{\footnotesize Merge overlays, infer flags}}};
\node[stage, right=1.8cm of manifest] (nef) {\shortstack{4. NEF Writer\\{\footnotesize Script + metadata}}};
\draw[->, thick] (frontend) -- node[above]{IR} (translator);
\draw[->, thick] (translator) -- node[above]{Script + metadata} (manifest);
\draw[->, thick] (manifest) -- node[above]{Manifest JSON} (nef);
\node[below=1.3cm of translator, align=center, font=\footnotesize] (helpers) {Runtime helper synthesis\\(memory, table, data)};
\draw[->, thick, dashed] (helpers) -- (translator);
\node[below=1.3cm of manifest, align=center, font=\footnotesize] (overlays) {Embedded + CLI overlays};
\draw[->, thick, dashed] (overlays) -- (manifest);
\caption{High-level Wasm $\rightarrow$ NeoVM translation pipeline. Solid arrows show the main data path, dashed arrows highlight auxiliary inputs.}
\label{fig:pipeline}
\end{tikzpicture}
\end{figure}

\section{Configuration Interfaces}
\subsection{Library Entry Point}
\begin{itemize}[nosep]
    \item \texttt{translate\_with\_config(bytes, TranslationConfig)} is the normative API. Required fields: contract name plus an optional \texttt{ManifestOverlay} (value + label) merged during manifest assembly.
    \item \texttt{translate\_module(bytes, name)} is a convenience wrapper equivalent to calling \texttt{translate\_with\_config} without overlays.
\end{itemize}

\subsection{CLI}
\begin{itemize}[nosep]
    \item \texttt{--input}: Wasm module path (required).
    \item \texttt{--nef}, \texttt{--manifest}: Output locations (default: derived from the input basename).
    \item \texttt{--name}: Contract name stored in the manifest (defaults to \texttt{Contract}).
    \item \texttt{--manifest-overlay}: External overlay JSON merged after embedded overlays.
    \item \texttt{--source-url}: Overrides the source URL recorded in NEF metadata.
    \item \texttt{--compare-manifest}: Reference manifest path; translation fails after printing a diff when the generated manifest diverges from this file.
\end{itemize}

\section{Wasm Module Acceptance Rules}
\subsection{Types and Functions}
\begin{itemize}[nosep]
    \item Function parameters and results are restricted to \texttt{i32} and \texttt{i64}. Multi-value returns are rejected, as are typed \texttt{select} or \texttt{call\_indirect} signatures that yield more than one result.
    \item Local and global initialiser expressions must fold to constants using \texttt{i32.const}, \texttt{i64.const}, or other constant expressions allowed by the Wasm core spec (no \texttt{global.get} chains).
    \item Each Wasm function references a type index. The translator records the mapping in \texttt{ModuleTypes} and aborts if a function references an unknown signature.
\end{itemize}

\subsection{Exports}
\begin{itemize}[nosep]
    \item Only functions may be exported. Duplicate export aliases are permitted and each alias maps to the same function index.
    \item All export aliases must reference an imported or defined function that meets the translation constraints. Unprocessed aliases cause translation to fail.
\end{itemize}

\subsection{Imports}
\begin{itemize}[nosep]
    \item Only function imports are permitted. Importing memories, tables, or globals is rejected.
    \item Each import records its module/name/type index. If the import is later exported, the translator generates a NeoVM stub so that the manifest still references an executable entry point.
    \item At most one start section is accepted. The referenced function (imported or defined) must not take parameters nor return values.
\end{itemize}

\subsection{Memories}
\begin{itemize}[nosep]
    \item Exactly one 32-bit, non-shared memory is supported. \texttt{memory64}, multiple memories, and shared memories are rejected.
    \item All memory indices must be zero for instructions and data segments. Offsets must be non-negative and fit into 32-bit space.
    \item Passive data segments require a defined memory section. Active segments targeting a non-zero memory index are rejected.
\end{itemize}

\subsection{Data Segments}
\begin{itemize}[nosep]
    \item Active segments must specify an \texttt{i32.const} or \texttt{i64.const} offset expression and fit within the declared initial memory size.
    \item Passive segments are cached in static slots together with a drop flag. \texttt{data.drop} may only execute once per segment, matching Wasm semantics.
\end{itemize}

\subsection{Tables}
\begin{itemize}[nosep]
    \item Tables must store \texttt{funcref} elements. \texttt{table64}, shared tables, and tables with other element types are rejected.
    \item Only declared tables (index~0) are supported. Table imports are rejected. Declared maximum limits are honoured by \texttt{table.grow}.
\end{itemize}

\subsection{Element Segments}
\begin{itemize}[nosep]
    \item Active segments specify a table index (default 0) and a constant offset expression. Passive segments are cached similarly to data segments. Declared-only segments are rejected.
    \item Segment payloads may be function indices or expressions made of \texttt{ref.null funcref} / \texttt{ref.func}. Other heap types and instructions are rejected.
    \item \texttt{elem.drop} may only be executed once per passive element segment.
\end{itemize}

\subsection{Globals}
\begin{itemize}[nosep]
    \item Global value types are limited to \texttt{i32}/\texttt{i64}. Immutable globals retain a compile-time literal until written via \texttt{global.set}.
    \item Global initialisers must fold to constants with no side-effects; the translator stores the initial value in a static slot.
\end{itemize}

\subsection{Start Section}
\begin{itemize}[nosep]
    \item At most one start function is permitted. It must not take parameters nor return values, regardless of whether it refers to an import or a defined function.
    \item The start function executes exactly once immediately before the first exported function (or runtime helper) requires access to memory, globals, or tables.
\end{itemize}

\subsection{Custom Sections}
\begin{itemize}[nosep]
    \item Sections named \texttt{.custom\_section.neo.manifest} (or suffixed variants) are parsed as concatenated JSON fragments and merged into the manifest.
    \item Sections named \texttt{.custom\_section.neo.methodtokens} provide method tokens and an optional \texttt{source}/\texttt{nefSource} string. Other custom sections are ignored.
\end{itemize}

\subsection{Reserved Import Modules}
\begin{itemize}[nosep]
    \item \texttt{syscall}: names must match Neo interop descriptors (for example, \texttt{System.Storage.Get}). Each import lowers to \texttt{SYSCALL <hash>} and registers storage usage for manifest feature inference.
    \item \texttt{neo}: DevPack-friendly aliases (\texttt{storage\_read}, \texttt{notify}, \texttt{call\_contract}, etc.) resolve to canonical descriptors before emitting \texttt{SYSCALL}.
    \item \texttt{opcode}: imports invoke raw NeoVM opcodes. The helper \texttt{opcode::RAW} and \texttt{opcode::RAW4} append literal bytes for advanced patterns.
    \item \texttt{env}: limited to C runtime memory utilities (\texttt{memcpy}/\texttt{memmove}/\texttt{memset} and their builtin spellings). Each takes three \texttt{i32} parameters and optionally returns the destination pointer.
\end{itemize}

\section{Compatibility Matrix}
\begin{table}[htbp]
\centering
\begin{tabular}{@{}p{3.6cm}p{9.4cm}@{}}
\toprule
Area & Status / Notes \\
\midrule
Linear memory & Single 32-bit memory, index~0, non-shared. Passive data segments require a defined memory. Helpers perform bounds checks and little-endian conversions. \\
Tables and references & Only \texttt{funcref} tables. Table helpers cover get/set/size/grow/fill/copy/init/drop and use \texttt{-1} as the null sentinel. \\
Imports & Only functions may be imported. Reserved modules (syscall, neo, opcode, env) provide dedicated lowering; importing memories/tables/globals is rejected. \\
Globals & \texttt{i32}/\texttt{i64} only. Immutable globals are constant-folded. Mutable globals lose literals after \texttt{global.set}. \\
Function signatures & \texttt{i32}/\texttt{i64} parameters, at most one result, and no multi-value typed \texttt{select}. \texttt{call\_indirect} signatures must match \texttt{funcref} table entries exactly. \\
Numeric features & Integer arithmetic, comparisons, conversions, bit counts, and shifts are supported. Floating-point, SIMD, atomics, threads, and reference types other than \texttt{funcref} are rejected. \\
\bottomrule
\end{tabular}
\end{table}

\section{Instruction Lowering}
Each operator lowers to NeoVM opcodes while preserving Wasm stack discipline. The tables below enumerate all supported operators, the emitted NeoVM instructions, and noteworthy behaviours.

\subsection{Stack, Local, and Selection Operators}
\begin{longtable}{@{}p{4.7cm}p{3.8cm}p{6.5cm}@{}}
\caption{Stack, local, and selection operators}\\
\toprule
Wasm operator(s) & NeoVM emission & Notes \\
\midrule
\endfirsthead
\multicolumn{3}{l}{\textit{Stack, local, and selection operators (continued)}}\\
\toprule
Wasm operator(s) & NeoVM emission & Notes \\
\midrule
\endhead
\midrule
\multicolumn{3}{r}{\textit{Continued on next page}}\\
\bottomrule
\endfoot
\bottomrule
\endlastfoot
\texttt{nop} & -- (elided) & Instruction is dropped; constant trackers ignore it.
\tabularnewline
\texttt{drop} & \texttt{DROP} or literal truncation & If the dropped value is a pending literal, its bytes are truncated; otherwise \texttt{DROP} is emitted.
\tabularnewline
\texttt{i32.const}/\texttt{i64.const} & Smallest \texttt{PUSHINT*} & Literals are tracked so that subsequent \texttt{drop}/\texttt{select} operations can erase them.
\tabularnewline
\texttt{local.get} & \texttt{LDARG*}/\texttt{LDLOC*} & Parameters use \texttt{LDARG*}; locals use \texttt{LDLOC*} after being allocated in the function prolog.
\tabularnewline
\texttt{local.set} & \texttt{STARG*}/\texttt{STLOC*} & Value is stored into the appropriate slot. Constant tracking is updated accordingly.
\tabularnewline
\texttt{local.tee} & \texttt{STARG*}/\texttt{STLOC*} + reload & Stores the value and reloads it immediately, preserving stack discipline.
\tabularnewline
\texttt{global.get} & \texttt{LDSFLD*} or literal & Immutable globals keep their literal and emit \texttt{PUSHINT*}. Mutable globals load from their static slot after ensuring the runtime has been initialised.
\tabularnewline
\texttt{global.set} & \texttt{STSFLD*} & Mutability is enforced. The runtime init helper executes first, then the new value is written to the assigned slot.
\tabularnewline
\texttt{select}, \texttt{typed.select}, \texttt{typed.select\_multi} & \texttt{JMPIFNOT\_L}, \texttt{NIP} & The translator emits a conditional branch that drops the unselected value. Typed variants validate that the result arity is \textbf{one} and the type is \texttt{i32}/\texttt{i64}.
\tabularnewline
\end{longtable}

\subsection{Control Flow and Calls}
\begin{longtable}{@{}p{4.7cm}p{3.8cm}p{6.5cm}@{}}
\caption{Control flow and call operators}\\
\toprule
Wasm operator(s) & NeoVM emission & Notes \\
\midrule
\endfirsthead
\multicolumn{3}{l}{\textit{Control flow and call operators (continued)}}\\
\toprule
Wasm operator(s) & NeoVM emission & Notes \\
\midrule
\endhead
\midrule
\multicolumn{3}{r}{\textit{Continued on next page}}\\
\bottomrule
\endfoot
\bottomrule
\endlastfoot
\texttt{block}, \texttt{loop}, \texttt{if}, \texttt{else}, \texttt{end} & \texttt{JMP\_L}/\texttt{JMPIF\_L} fixups & Control frames track stack height and expected results. Branch targets patch deferred \texttt{JMP\_L} or \texttt{JMPIF\_L} immediates.
\tabularnewline
\texttt{br}, \texttt{br\_if} & \texttt{JMP\_L}/\texttt{JMPIF\_L} & \texttt{br} emits an unconditional jump to the recorded label. \texttt{br\_if} branches when the condition is non-zero.
\tabularnewline
\texttt{br\_table} & Cascaded \texttt{JMPIF\_L} & The selector is duplicated and compared against each target depth. A default label handles the fall-through case.
\tabularnewline
\texttt{return} & \texttt{RET} & Emits a return opcode, clears the abstract value stack, and prevents duplicate \texttt{RET} emission in the epilogue.
\tabularnewline
\texttt{call} & \texttt{CALL\_L} or import bridge & Defined functions emit \texttt{CALL\_L} fix-ups via the \texttt{FunctionRegistry}. Imports lower to \texttt{SYSCALL}, raw opcodes, or runtime helpers (\texttt{env} shims) depending on the module.
\tabularnewline
\texttt{call\_indirect} & Table helper + dispatch & Validates the table index, traps on null \texttt{funcref} entries, then matches the function index against the requested signature before emitting an import or defined call.
\tabularnewline
\texttt{unreachable} & \texttt{ABORT} & Forces a NeoVM fault and clears the abstract stack to maintain invariants.
\tabularnewline
\end{longtable}

\subsection{Integer Arithmetic, Comparisons, and Conversions}
\begin{longtable}{@{}p{4.7cm}p{3.8cm}p{6.5cm}@{}}
\caption{Numeric operators}\\
\toprule
Wasm operator(s) & NeoVM emission & Notes \\
\midrule
\endfirsthead
\multicolumn{3}{l}{\textit{Numeric operators (continued)}}\\
\toprule
Wasm operator(s) & NeoVM emission & Notes \\
\midrule
\endhead
\midrule
\multicolumn{3}{r}{\textit{Continued on next page}}\\
\bottomrule
\endfoot
\bottomrule
\endlastfoot
\texttt{i32.add}, \texttt{i64.add}, \texttt{i32.sub}, \texttt{i64.sub} & \texttt{ADD}/\texttt{SUB} & Operands remain full-width signed integers. Constant pairs fold at translation time.
\tabularnewline
\texttt{i32.mul}, \texttt{i64.mul} & \texttt{MUL} & Multiplication reuses NeoVM's full-width integer multiplier; overflow follows NeoVM semantics.
\tabularnewline
\texttt{i32.and}, \texttt{i64.and}, \texttt{i32.or}, \texttt{i64.or}, \texttt{i32.xor}, \texttt{i64.xor} & \texttt{AND}/\texttt{OR}/\texttt{XOR} & Bitwise ops operate on the full integer width.
\tabularnewline
\texttt{i32.shl}, \texttt{i64.shl}, \texttt{i32.shr\_s}, \texttt{i64.shr\_s}, \texttt{i32.shr\_u}, \texttt{i64.shr\_u} & \texttt{SHL}/\texttt{SHR} & Shift counts are masked to the operand width. Unsigned shifts mask the value before invoking \texttt{SHR}.
\tabularnewline
\texttt{i32.rotl}, \texttt{i64.rotl}, \texttt{i32.rotr}, \texttt{i64.rotr} & Stack shuffles + \texttt{SHL}/\texttt{SHR} & Rotation counts are masked; large rotations wrap naturally.
\tabularnewline
\texttt{i32.div\_s}, \texttt{i64.div\_s} & \texttt{DIV} & Signed division delegates to NeoVM. Constant operands are folded unless dividing by zero or the \texttt{INT\_MIN / -1} edge case.
\tabularnewline
\texttt{i32.div\_u}, \texttt{i64.div\_u} & Mask + \texttt{DIV} & Operands are zero-extended prior to division to reproduce unsigned semantics.
\tabularnewline
\texttt{i32.rem\_s}, \texttt{i64.rem\_s} & \texttt{MOD} & Signed remainder uses NeoVM's \texttt{MOD}. Division-by-zero checks happen via the underlying VM fault.
\tabularnewline
\texttt{i32.rem\_u}, \texttt{i64.rem\_u} & Mask + \texttt{MOD} & Operands are zero-extended before applying \texttt{MOD}.
\tabularnewline
\texttt{i32.clz}, \texttt{i32.ctz}, \texttt{i32.popcnt}, \texttt{i64.clz}, \texttt{i64.ctz}, \texttt{i64.popcnt} & Helper routines & Constant inputs fold; dynamic inputs call a lazily-emitted helper specialised per bit width.
\tabularnewline
\texttt{i32.eq}, \texttt{i64.eq}, \texttt{i32.ne}, \texttt{i64.ne} & \texttt{EQUAL}/\texttt{NOTEQUAL} & Returns 1 for true and 0 otherwise.
\tabularnewline
\texttt{i32.lt\_s}, \texttt{i64.lt\_s}, \texttt{i32.le\_s}, \texttt{i64.le\_s}, \texttt{i32.gt\_s}, \texttt{i64.gt\_s}, \texttt{i32.ge\_s}, \texttt{i64.ge\_s} & Comparison helpers & Signed comparisons reuse NeoVM relational operators.
\tabularnewline
\texttt{i32.lt\_u}, \texttt{i64.lt\_u}, \texttt{i32.le\_u}, \texttt{i64.le\_u}, \texttt{i32.gt\_u}, \texttt{i64.gt\_u}, \texttt{i32.ge\_u}, \texttt{i64.ge\_u} & Mask + comparisons & Operands are zero-extended before invoking the signed comparison to emulate unsigned semantics.
\tabularnewline
\texttt{i32.eqz}, \texttt{i64.eqz} & \texttt{PUSH0} + \texttt{EQUAL} & Literals fold; dynamic values compare against zero at runtime.
\tabularnewline
\texttt{i32.wrap\_i64}, \texttt{i64.extend\_i32\_u}, \texttt{i64.extend\_i32\_s} & Mask/shift helpers & \texttt{i32.wrap\_i64} masks the lower 32 bits. Extensions reuse \texttt{emit\_sign\_extend} or \texttt{emit\_zero\_extend} helpers.
\tabularnewline
\texttt{i32.extend8\_s}, \texttt{i32.extend16\_s} & Sign-extension helper & Shifts and masks reconstitute the signed value within 32 bits.
\tabularnewline
\texttt{i64.extend8\_s}, \texttt{i64.extend16\_s}, \texttt{i64.extend32\_s} & Sign-extension helper & Equivalent to the 32-bit case but for wider targets.
\tabularnewline
\end{longtable}

\subsection{Memory Operators}
\begin{longtable}{@{}p{4.7cm}p{3.8cm}p{6.5cm}@{}}
\caption{Memory operators}\\
\toprule
Wasm operator(s) & NeoVM emission & Notes \\
\midrule
\endfirsthead
\multicolumn{3}{l}{\textit{Memory operators (continued)}}\\
\toprule
Wasm operator(s) & NeoVM emission & Notes \\
\midrule
\endhead
\midrule
\multicolumn{3}{r}{\textit{Continued on next page}}\\
\bottomrule
\endfoot
\bottomrule
\endlastfoot
Loads: \texttt{i32.load}, \texttt{i64.load}, \texttt{i32.load8\_s}, \texttt{i32.load8\_u}, \texttt{i32.load16\_s}, \texttt{i32.load16\_u}, \texttt{i64.load8\_s}, \texttt{i64.load8\_u}, \texttt{i64.load16\_s}, \texttt{i64.load16\_u}, \texttt{i64.load32\_s}, \texttt{i64.load32\_u} & Helper + \texttt{SUBSTR}/\texttt{CONVERT} & The runtime init helper executes first (allocating the buffer, applying data segments, running \texttt{start}). Addresses are bounds-checked and values are loaded little-endian, zero- or sign-extended per opcode.
\tabularnewline
Stores: \texttt{i32.store}, \texttt{i64.store}, \texttt{i32.store8}, \texttt{i32.store16}, \texttt{i64.store8}, \texttt{i64.store16}, \texttt{i64.store32} & Helper + \texttt{SETITEM} & Validates addresses and writes bytes little-endian via repeated \texttt{SETITEM} calls.
\tabularnewline
\texttt{memory.size} & \texttt{LDSFLD2} & Returns the current page count (32-bit pages). Requires prior memory definition.
\tabularnewline
\texttt{memory.grow} & Helper routine & Masks the delta to 32 bits, enforces the declared maximum (or returns \texttt{-1}), allocates a new buffer when necessary, copies existing bytes, and returns the previous page count.
\tabularnewline
\texttt{memory.fill}, \texttt{memory.copy}, \texttt{memory.init} & Helper routines & Validate offsets/lengths, operate on the buffer in-place (with overlap-safe copies), and reference passive data slots for \texttt{memory.init}. Out-of-bounds or negative arguments trap via \texttt{ABORT}.
\tabularnewline
\texttt{data.drop} & Helper routine & Each passive segment has a drop flag. Dropping more than once triggers \texttt{ABORT}.
\tabularnewline
Imported \texttt{env::memcpy}, \texttt{env::memmove}, \texttt{env::memset} & Helper stubs & Resolved imports call the same runtime helpers as \texttt{memory.copy} and \texttt{memory.fill}, keeping \texttt{-nostdlib} builds self-contained.
\tabularnewline
\end{longtable}

\subsection{Table and Reference Operators}
\begin{longtable}{@{}p{4.7cm}p{3.8cm}p{6.5cm}@{}}
\caption{Table and reference operators}\\
\toprule
Wasm operator(s) & NeoVM emission & Notes \\
\midrule
\endfirsthead
\multicolumn{3}{l}{\textit{Table and reference operators (continued)}}\\
\toprule
Wasm operator(s) & NeoVM emission & Notes \\
\midrule
\endhead
\midrule
\multicolumn{3}{r}{\textit{Continued on next page}}\\
\bottomrule
\endfoot
\bottomrule
\endlastfoot
\texttt{table.get}, \texttt{table.set}, \texttt{table.size} & Helper routines on static arrays & Tables live in static slots as NeoVM arrays of signed integers (\texttt{-1} = null). Helpers perform bounds checks before reading/writing entries or returning the array length.
\tabularnewline
\texttt{table.grow} & Helper routine & Respects the declared maximum. New entries are initialised to \texttt{-1}. Returns the previous table length or \texttt{-1} on failure.
\tabularnewline
\texttt{table.fill}, \texttt{table.copy} & Helper routines & Validate lengths and ranges, copy entries via temporary arrays, and abort on overlap violations or missing tables.
\tabularnewline
\texttt{table.init}, \texttt{elem.drop} & Helper routines with passive caches & Passive element segments are cached (value array + drop flag). \texttt{table.init} copies values into the target slice; \texttt{elem.drop} flips the drop flag and aborts when invoked twice.
\tabularnewline
\texttt{ref.null}, \texttt{ref.is\_null} & Push \texttt{-1}/compare to \texttt{-1} & Only \texttt{funcref} heap types are accepted. Null references push the sentinel value.
\tabularnewline
\texttt{ref.func} & \texttt{PUSHINT*} & Encodes the function index as an integer after validating that the function exists.
\tabularnewline
\texttt{ref.as\_non\_null} & Null check + \texttt{ABORT} & Duplicates the reference, compares against the null sentinel, and traps if null.
\tabularnewline
\texttt{ref.eq} & \texttt{EQUAL} & Performs integer equality against the sentinel-encoded references.
\tabularnewline
\end{longtable}

\subsection{Unsupported Features}
The translator rejects:
\begin{itemize}[nosep]
    \item Floating-point instructions, SIMD, atomics, threads, and reference types other than \texttt{funcref}.
    \item Multi-value returns, multi-value typed \texttt{select}, \texttt{memory64}, multiple memories, shared memories/tables, and table imports.
    \item Declared-only element segments, \texttt{call\_indirect} signatures with unsupported types, and repeated \texttt{data.drop}/\texttt{elem.drop} invocations.
\end{itemize}

\section{Runtime Helpers}
\subsection{Linear Memory Layout}
\begin{table}[htbp]
\centering
\begin{tabular}{@{}c p{10cm}@{}}
\toprule
Slot & Purpose \\
\midrule
0 & NeoVM \texttt{BUFFER} handle backing Wasm linear memory. \\
1 & Current byte length of the buffer. \\
2 & Current page count (32-bit pages). \\
3 & Declared maximum page count or \texttt{-1} when unbounded. \\
4 & Initialisation flag (0 before the init helper runs, 1 afterwards). \\
\bottomrule
\end{tabular}
\caption{Base static slot layout}
\end{table}
Memory initialisation allocates static slots via \texttt{INITSSLOT}, creates the buffer, records the declared limits, and stores \texttt{0} in the init flag. Additional slots are allocated sequentially for globals, tables, passive data (two slots per segment: bytes + drop flag), passive elements (two slots per segment: value array + drop flag), and an optional start guard bit; see Figure~\ref{fig:memory-slots}.

Every instruction that may touch runtime state begins with \texttt{emit\_memory\_init\_call}, which loads slot~4. If zero, the helper allocates memory, applies active data segments, initialises tables/elements, runs the start function exactly once, and sets the flag to 1. Subsequent calls skip the helper entirely.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[every node/.style={font=\small}, node distance=0cm]
\tikzstyle{slot}=[draw, minimum width=7cm, minimum height=0.9cm, anchor=west, fill=gray!5]
\node[slot] (slot0) {Slot 0 \textbf{Buffer handle} (\texttt{LDSFLD0})};
\node[slot, below=0cm of slot0] (slot1) {Slot 1 \textbf{Buffer length} (bytes)};
\node[slot, below=0cm of slot1] (slot2) {Slot 2 \textbf{Page count} (32-bit pages)};
\node[slot, below=0cm of slot2] (slot3) {Slot 3 \textbf{Max pages} (\texttt{-1} = unbounded)};
\node[slot, below=0cm of slot3, fill=gray!15] (slot4) {Slot 4 \textbf{Init flag} (0 prior to helper, 1 afterwards)};
\draw[decorate, decoration={brace, amplitude=6pt}] ([xshift=-0.4cm]slot0.north west) -- node[left=0.2cm, align=right]{\footnotesize Base memory slots\\(always reserved)} ([xshift=-0.4cm]slot4.south west);
\node[below=0.2cm of slot4, align=center, font=\footnotesize] {Slots for globals, tables, passive data/element caches, and the optional start guard extend the layout further.};
\caption{Static slot layout reserved by the memory initialiser. Figure~\ref{fig:pipeline} shows where the helper sits in the overall pipeline.}
\label{fig:memory-slots}
\end{figure}

\subsection{Memory Helpers}
Load/store helpers take an address, ensure it is non-negative, check that \texttt{address + width} does not exceed the current length, and operate on the buffer using \texttt{SUBSTR}, \texttt{CONVERT}, and \texttt{SETITEM}. \texttt{memory.grow} masks the delta, enforces the maximum, and copies the old buffer into the new allocation via \texttt{MEMCPY}. \texttt{memory.fill}, \texttt{memory.copy}, and \texttt{memory.init} allocate local slots, validate lengths, and call the corresponding NeoVM primitives. Passive data segments keep their bytes plus a drop flag; \texttt{memory.init} requires the segment to be present and \texttt{data.drop} flips the flag (aborting when invoked twice). Imported \texttt{env::memcpy}, \texttt{env::memmove}, and \texttt{env::memset} invoke the same helpers, ensuring \texttt{-nostdlib -fno-builtin} contracts remain pure Wasm.

\subsection{Table Helpers}
Each Wasm table becomes a NeoVM array stored in a static slot. Entries encode function indices as signed integers with \texttt{-1} representing null. Helpers exist for \texttt{table.get}, \texttt{table.set}, \texttt{table.size}, \texttt{table.grow} (respecting maxima), \texttt{table.fill}, \texttt{table.copy}, \texttt{table.init}, and \texttt{elem.drop}. Passive element segments store two slots (value array + drop flag); \texttt{table.init} copies values into the destination slice after verifying both the table bounds and the segment drop flag. \texttt{elem.drop} writes \texttt{1} into the flag and aborts if repeated.

\subsection{Start Execution Guard}
If a start section is present, the runtime helper stores an additional guard slot. Every translated function (except the start function itself) emits \texttt{emit\_memory\_init\_call} at entry so the start function runs before contract code touches any global state. Imported start functions call the respective import bridge; defined start functions reuse their translated offset. When the start function is the only exported entry, the translator appends a stub that first runs the init helper and then dispatches to the start body, updating the exported offset accordingly. Init failures propagate (the helper traps instead of swallowing errors), and the guard still ensures the start code executes exactly once.

\section{Custom Sections and Metadata}
\subsection{Manifest Overlays}
\texttt{neo.manifest} custom sections contain one or more concatenated JSON fragments. Each fragment is merged into the baseline manifest using deep merges: nested objects merge recursively, arrays are concatenated, and duplicate strings/events/permissions/trusts are deduplicated. The translator also scans \texttt{abi.methods[*].safe} flags inside the overlay so that safe designations survive later merges.

\subsection{Method Tokens and Source Hints}
\texttt{neo.methodtokens} custom sections contain either JSON arrays or objects. Each fragment yields \texttt{MethodToken} entries (contract hash, method name, parameter count, return flag, call flags) plus an optional \texttt{source}/\texttt{nefSource} override. The translator also heuristically infers method tokens by scanning the generated script for \texttt{System.Contract.Call} sequences (contract hash literal, method literal, call flags, argument array). Duplicate tokens are removed by hashing the tuple of (hash, method, parameter count, return flag, call flags). When multiple fragments provide a source URL, the earliest one wins unless the CLI override is supplied.

\section{Manifest Generation}
\subsection{Baseline Manifest}
Each exported alias becomes an \texttt{abi.methods} entry named after the alias. Parameters are rendered as \texttt{Integer} with names \texttt{arg0}, \texttt{arg1}, etc. Return values translate to \texttt{Integer} or \texttt{Void}. Default metadata includes empty \texttt{groups}, \texttt{supportedstandards}, \texttt{permissions}, \texttt{trusts}, and an \texttt{extra} block with author/description placeholders.

\subsection{Overlay Application and Parity}
Embedded \texttt{neo.manifest} fragments are merged first, then the CLI-provided overlay (if any). The \texttt{ManifestBuilder} keeps the baseline method set and records the label of the last overlay file so that diagnostics can point to the offending source. After merging, safe flags propagate across duplicate entries, features are inferred automatically, and the builder aborts if overlays introduce or remove ABI methods \emph{or} mutate existing signatures/offsets. The CLI's \texttt{--compare-manifest} flag compares the rendered manifest against a reference file and prints a diff when they diverge.

\subsection{Feature Inference}
The \texttt{FeatureTracker} inspects imported syscalls and exported method names. Invoking \texttt{System.Storage.*} sets \texttt{features.storage=true}. Exporting \texttt{onPayment}, \texttt{onNEP17Payment}, or \texttt{onNEP11Payment} toggles \texttt{features.payable=true}. The manifest rarely needs to toggle these flags manually.

\section{NEF Emission}
\subsection{Metadata}
\texttt{write\_nef\_with\_metadata} produces NEF v3 files with the layout:
\begin{itemize}[nosep]
    \item Magic number (\texttt{NEF3}) and a 64-byte compiler string (\texttt{neo-llvm wasm-neovm <version>}).
    \item Source URL encoded as a VarString (max 256 bytes) sourced from custom sections, overlays, or \texttt{--source-url}.
    \item Reserved byte (always zero).
    \item Method-token block encoded as a VarUint count followed by tokens (contract hash, method name, parameter count, return flag, call flags) and a 2-byte terminator.
    \item Script payload as a VarBytes field and a trailing double-SHA256 checksum.
\end{itemize}
Method tokens come from overlay metadata, \texttt{neo.methodtokens} sections, inferred call sequences, and manifest \texttt{extra.nefMethodTokens}. Duplicates are removed before writing. NEF emission guarantees the manifest and script stay in sync; failures abort before writing either artefact.

Method names in tokens are limited to 32 bytes per the NEF schema. Oversized names (e.g., long syscall identifiers) are rejected at parse time, and the inference pass skips any detected sequences whose method literal exceeds the limit.

\section{Errors and Diagnostics}
Translation aborts with descriptive diagnostics on:
\begin{itemize}[nosep]
    \item Unsupported Wasm features (floating point, multi-memory, reference types other than \texttt{funcref}, declared element segments, etc.).
    \item ABI mismatches introduced by overlays or \texttt{--compare-manifest} differences.
    \item Out-of-bounds memory/table accesses discovered while synthesising runtime helpers, or repeated \texttt{data.drop}/\texttt{elem.drop} invocations.
    \item Invalid custom-section payloads (non-JSON \texttt{neo.manifest}/\texttt{neo.methodtokens}), env imports with the wrong signature, or start sections referencing non-functions.
\end{itemize}

\section{Conformance}
An implementation conforms to this specification if it:
\begin{itemize}[nosep]
    \item Accepts any Wasm module satisfying the above rules and emits a NeoVM script whose observable behaviour matches the instruction mappings.
    \item Produces NEF and manifest artefacts that reflect the translated ABI, overlays, inferred metadata, and method tokens while preserving safe flags and feature inference.
    \item Rejects inputs that violate acceptance rules, manifest parity, or helper invariants with actionable diagnostics, and keeps the script/manifest/NEF outputs mutually consistent.
\end{itemize}

\end{document}
